---
title: "Introduction to R and Rstudio"
subtitle: "Session 4: Data Wrangling: Part B"
output: html_notebook
---

First use a command to clear all the environment variables
```{r}
rm(list=ls())
```


##5 Summarising with Tidyverse
```{r}
# load tidyverse
library(tidyverse)
```


Now that we have our data organised, we can summarise it.

First we must read in the csv file from part 1
```{r}
b5.en_two  <-  read.csv("b5_en_two.csv")
```

First we will summarise all of the data
Then summarise by group

###Summarise all

first take a look at the column names
```{r}
b5.en_two$ID  <-  factor(seq(1:nrow(b5.en_two))) # create an ID variable
names(b5.en_two)
```

So we want to keep X, Eng, Name_en, Age and Gender and all the items and remove the rest of the variable.
We could i apply a NULL value to each column, or ii we could use the not `-` operator next to the `c()` function inside the `select()` function: 
Lets do number ii:
```{r}
b5.en_two %>%
  select(-c(race, engnat, hand, source)) %>%
  select(ID, age, gender, country,ENG, Name_en, everything()) ->> b5.tdy1

b5.tdy1$gender <- factor(b5.tdy1$gender)
b5.tdy1$ENG    <- factor(b5.tdy1$ENG)  
b5.tdy1$X.1 <- NULL # Remove the superfluos variable
b5.tdy1
```

We can recode the gender Factor to Representthe gender names (Values in the code book)
```{r}

b5.tdy1%>%
  mutate(gender, gender  = fct_recode(gender, 
                                      "None"   = "0",
                                      "Male"   = "1", 
                                      "Female" = "2",
                                      "Other"  = "3")) ->> b5.tdy1
```


Now we can view the new df
```{r}
b5.tdy1
```
Next we want to transform the data into a long format data frame, keeping ID, age, gender and country constant.
If we look a the column description, we can see that some columns are integer and some are numeric.

So what we have to do first is transform all of the items into factor variables and just check their levels
```{r}
b5.tdy1 %>%
  select(E1:O10) -> items
str(items)
```
So all columns are integers

Lets just take a look at the levels of the items, to ensure that there arent any values present that should not be
```{r}
iname <- names(items) #  create a list of variable names

items %>%
  mutate_all(funs(as.character(.)),iname) %>%
  mutate_if(is.character, as.factor) %>% 
  sapply(levels)
```

There is the correct number of levels (they should all be 1-5). So now we should convert all response data to one type:

```{r}
b5.tdy1 %>%
  mutate_at(funs(as.numeric), .vars=iname) ->> b5.tdy2

b5.tdy2
```

So now the b5.tdy2 data frame is ready for mutation. If we transform the data frame to a tidy df, with a factor level for each item, then we will be able to summarise responses to each item
```{r}
library(reshape2)
```

We can use melt to transform
```{r}
b5.tdy2 %>%
  melt(id.vars = c("ID", "age", "gender", "country", "ENG", "Name_en"),  
       variable.name="Item", 
       value.name="Response") ->> b5.tdy3
b5.tdy3
```

Now we can go one step further and create a factor/Dimension label for to be attached using a regex and ifelse function
```{r}
b5.tdy3$Dimension <- b5.tdy3$Item
b5.tdy3$Dimension <- sub("^(.).*(\\d).*$", "\\1\\", b5.tdy3$Dimension) 
b5.tdy3$Dimension <-  factor(b5.tdy3$Dimension)

# Order the factor levels
ord <- c("O", "C", "E", "A", "N")
b5.tdy3$Dimension <- factor(b5.tdy3$Dimension,levels=ord)

b5.tdy3 %>%
  select(ID, age, gender, country,ENG, Name_en, Dimension, everything()) ->> b5.tdy4
```

And now we can view our new data frame
```{r}
b5.tdy4
```


And now we can summarise our data by using factor variables in the group-by argument
```{r}
# By Dimension summary
b5.tdy4%>%
  group_by(Dimension) %>%
  summarise(Mean = mean(Response), 
            StDev = sd(Response), 
            Count = n(), 
            StdErr = StDev/sqrt(Count))
```

```{r}
# By gender Summary
b5.tdy4 %>%
  group_by(Dimension, gender) %>%
  summarise(Mean = mean(Response), 
            StDev = sd(Response), 
            Count = n(), 
            StdErr = StDev/sqrt(Count))
```

